---
title: "Designing a 60 ft Loaded 80 m Dipole with loadedDipole"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Designing a 60 ft Loaded 80 m Dipole with loadedDipole}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(loadedDipole)
```

## Overview

This vignette walks through the design of a shortened 80 m dipole for
3.57 MHz FT8, using a total wire length of 60 ft and a pair of
available 66 µH loading coils.

We will:

1. Compute the recommended coil placement.
2. Visualise inductance versus coil placement.
3. Estimate radiation efficiency for the final design.

## Antenna and coil parameters

```{r}
frequency    <- 3.57     # MHz (FT8)
total_length <- 60       # ft
L_coil       <- 66       # microhenry per coil
```

## 1. Recommended coil placement

```{r}
rec <- recommend_coil_position(
  frequency    = frequency,
  total_length = total_length,
  inductance   = L_coil
)

rec
```

The output reports the recommended distance from the feedpoint to each coil
(per leg) and the fraction of the half-element length at which the coils
are placed.

## 2. Inductance versus placement

```{r, fig.width=8, fig.height=5}
plot_inductor_placement(
  frequency    = frequency,
  total_length = total_length
)
```

This plot shows the required inductance as a function of coil position
for the chosen frequency and total length. The recommended coil position
is indicated by a dashed vertical line.

## 3. Efficiency estimate

```{r}
eff <- estimate_efficiency(
  frequency       = frequency,
  total_length    = total_length,
  coil_inductance = L_coil,
  coil_position   = rec$coil_position,
  model           = "both"
)

eff
```

The resulting table reports approximate radiation resistance, loss
resistance and overall radiation efficiency for two simple models:

- `simple`: short-dipole radiation resistance plus coil loss.
- `segmented`: adds crude wire and ground loss and a tapered current model.

The estimates are intended for comparative design work and trade-off
studies rather than as a replacement for full NEC modelling.

## 4. Efficiency vs Coil Position (Realistic Sweep)

In this section we sweep **coil position** *and compute the required
inductance at each position*, then evaluate the efficiency. This gives a
realistic view of how coil placement affects performance.

```{r, fig.width=8, fig.height=5}
library(ggplot2)

wire_diameter = 0.065

# sweep from 10% to 90% of half-length
half_len <- total_length / 2
fractions <- seq(0.01, 0.99, by = 0.01)
coil_positions <- fractions * half_len

eff_list <- lapply(seq_along(coil_positions), function(i) {
  B <- coil_positions[i]

  # compute required inductance at this coil position
  L_req <- loading_coil_inductance(
    frequency,
    total_length = total_length,
    coil_position = B,
    wire_diameter = wire_diameter,
    metric = FALSE
  )

  # compute efficiency using segmented model
  eff <- estimate_efficiency(
    frequency       = frequency,
    total_length    = total_length,
    coil_inductance = L_req,
    coil_position   = B,
    model           = "segmented"
  )

  data.frame(
    coil_position = B,
    fraction_of_half = fractions[i],
    inductance = L_req,
    efficiency_loss = -10 * log10(eff$efficiency)

  )
})

eff_df <- do.call(rbind, eff_list)

# plot the result
ggplot(eff_df, aes(x = coil_position, y = efficiency_loss)) +
  geom_line(size = 1.1) +
  geom_vline(
    xintercept = rec$coil_position,
    linetype = "dashed",
    color = "red"
  ) +
  annotate(
    "text",
    x = rec$coil_position,
    y = max(eff_df$efficiency),
    label = sprintf("Recommended: %.2f ft", rec$coil_position),
    vjust = -0.5,
    color = "red"
  ) +
  labs(
    x = "Coil position from feedpoint (ft)",
    y = "Efficiency Loss (db)",
    title = "Efficiency vs Coil Position (Required Inductance Computed per Position)",
    subtitle = "Dashed line shows recommended coil placement"
  ) +
  theme_minimal()
```

This sweep computes:

- the *actual required inductance* at each B  
- the corresponding coil loss  
- the effect of shortening on radiation resistance  
- and thus the true efficiency curve  

## Optimizing coil inductance vs placement

A common design question is:

> *“If I move the coils inward, can I use a smaller inductance without
>  hurting efficiency?”*

`optimize_inductance_for_position()` allows us to sweep coil placements,
compute the required inductance, and optionally compute efficiency.

Here we analyze a 60 ft 80-meter dipole at 3.57 MHz:

```{r}

positions <- seq(8, 22, by = 0.5)   # ft from feedpoint

opt_df <- optimize_inductance_for_position(
  frequency       = 3.57,
  total_length    = 60,
  positions       = positions,
  efficiency      = TRUE,
  efficiency_model = "segmented"
)

opt_df
```

We can now visualize how inductance and efficiency vary with coil placement:

```{r, fig.width=8, fig.height=5}

ggplot(opt_df, aes(x = coil_position, y = inductance)) +
  geom_line(size = 1) +
  labs(
    title = "Required inductance vs coil position",
    x     = "Coil position from feedpoint (ft)",
    y     = "Inductance (µH)"
  )
```

```{r, fig.width=8, fig.height=5}
ggplot(opt_df, aes(x = coil_position, y = eff_segmented)) +
  geom_line(size = 1) +
  labs(
    title = "Efficiency vs coil position (segmented model)",
    x     = "Coil position from feedpoint (ft)",
    y     = "Efficiency (0–1)"
  )
```

This demonstrates that:

- Moving coils inward reduces required inductance dramatically  
- Efficiency stays nearly flat, as expected for a short 80-meter dipole  
- Practical positions (≈10–15 ft from the feedpoint) allow much smaller,
  lower-loss coils without meaningful loss of radiated power


## Designing a loaded dipole when only the total length is known

Often a builder only knows the available span (e.g., 60 ft) and wishes to
design the rest of the antenna with the constraint that the loading coils
must not exceed some value. Here we solve the problem for a 60 ft dipole
at 3.57 MHz, allowing coils of up to 66 µH.

```{r design-60ft, message=FALSE, fig.width=8, fig.height=4}
library(loadedDipole)

design <- design_loaded_dipole(
  frequency    = 3.57,
  total_length = 60,
  L_max        = 66
)

design$best
```

The output gives the recommended coil placement and inductance that:

1. Requires no more than 66 µH,  
2. Is within 0.1 dB of the best achievable efficiency, and  
3. Uses the smallest inductance within that range.

We can also visualize the entire design space:

```{r design-plot, fig.width=9, fig.height=4}
library(ggplot2)

df <- design$designs

ggplot(df, aes(x = coil_position, y = inductance)) +
  geom_line(color="steelblue", size=1) +
  geom_point(data = design$best, aes(x = coil_position, y = inductance),
             color="red", size=3) +
  labs(
    title = "Required inductance vs coil position (60 ft dipole, 3.57 MHz)",
    x = "Coil position (ft from feedpoint)",
    y = "Inductance (µH)"
  )
```

```{r design-efficiency-plot, fig.width=9, fig.height=4}
ggplot(df, aes(x = coil_position, y = loss_dB)) +
  geom_line(size=1) +
  geom_point(data = design$best, aes(x = coil_position, y = loss_dB),
             color="red", size=3) +
  labs(
    title = "Efficiency loss vs coil position (segmented model)",
    x = "Coil position (ft)",
    y = "Loss relative to perfect dipole (dB)"
  )
```


## Designing and Building the Loading Coils

Once the optimal coil inductance is found (e.g., ~33.5 uH at 8 ft from the
feedpoint), we can design the actual coils using PVC pipe and common
magnet wire. The function `design_pvc_coil()` computes the required turns,
coil geometry, and a ±2 turn table showing how inductance varies with turn
count.

In this example, we design a 33.5 uH coil on 1.5 inch Schedule 40 PVC with
AWG 14 wire (0.064") and 0.070" turn spacing:

```{r pvc-design, message=FALSE}
coil <- design_pvc_coil(
  inductance     = 33.5,
  pvc_size       = "1-1/2",
  wire_diameter  = 0.064,
  turn_spacing   = 0.070
)

coil
```

This gives:

- The computed turns (`coil$turns`),
- Winding length,
- Wire length,
- And a small table showing `turns +/- 2` so you can easily trim or add turns.

We can also visualize inductance vs turns for this PVC size:

```{r pvc-plot, fig.width=8, fig.height=4}
plot_coil_turns(
  pvc_size       = "1-1/2",
  wire_diameter  = 0.064,
  turn_spacing   = 0.070,
  N_range        = 10:70
)
```

In practice, wind one or two turns extra, measure the inductance using an
LCR meter or analyzer, and remove/spread turns until the value matches the
design target.


## Render a full HTML build report for a loaded dipole design

You can get a full report that gives tables and a figure for construction.

```{r full_report, eval = FALSE}
des <- design_loaded_dipole_full(frequency = 3.574, total_length = 60, L_max = 66, 
                                 pvc_size = "1-1/2", make_plot = FALSE 
                                  )
render_loaded_dipole_html(
  design = des,
  file = "loaded_dipole_report.html",
  include_plot = FALSE
)

```
Our output file will look like this:

Just print, grab your PVC,wire, soldering iron and insulators and go!
```{r full_reportx, echo = FALSE}
des <- design_loaded_dipole_full(frequency = 3.574, total_length = 60, L_max = 66, 
                                 pvc_size = "1-1/2", make_plot = FALSE
                                  )
htmltools::HTML(
 render_loaded_dipole_html(
  design = des,
  file = NULL,
  include_plot = TRUE
  )
)

```